Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2019-02-13 Manuscript/manuscript.Rmd
manuscript edits

Commonly used measures of model performance for binary classifiers include the area under the receiver operating curve (AUROC). The popularity surrounding the use of ROC curves is that they do not depend on a threshold. However, when the prevalence of an event is low--that is, a defined genomic region containing a TAD boundary--mean probabilities are biased towards the majority class. This results in the area of interest under an ROC curve to be compressed toward a small corner in the ROC space [@prati2011survey]. While not directly affecting the value of the AUROC, this can have severe consequences on the interpretability of the model's performance. A better threshold-free measure to use when dealing with imbalanced data is the area under the precision-recall curve (AUPRC). It has been shown that the use of AUPRC with imbalanced datasets yields more accurate and intuitive interpretation of model performance when compared to AUROC [@saito2015precision]. Likewise, metrics like accuracy and specificity are also problematic when evaluating model performance on imbalance data due to being highly influenced by the correct classification of the majority class. Better measures to consider are the precision and recall themselves. Both the precision and recall of a model include the true positives in the numerator of their respective ratios, thereby yielding more appropriate measures of a model's ability to accurately classify positive classes. Therefore, certain composite metrics like the F1-Score, Youden's Index, and Matthew's Correlation Coefficient (MCC) offer better interpretation, especially in the instance of imbalanced classes [@jeni2013facing]. The F1-Score can be interpreted as a weighted average of the precision and recall values, while Youden's index is interpreted as the arithmetic mean between sensitivity (recall) and specificity. MCC is, in essence, the correlation coefficient between the observed and predicted binary classifications. Each metric uses all 4 quadrants of the confusion matrix, thus providing more balanced measures in the case of imbalanced data [@chicco2017ten].

@@

To establish the data for modeling, the genome was binned into equally sized intervals defined by the data resolution for the corresponding contact matrix (Figure 1). For example, for domain data obtained from a 10 kb resolution contact matrix, the genome was binned into 10 kb intervals. To accommodate for the uncertainty in exact boundary location, the boundary points were flanked on both sides by the size of the resolution of the corresponding contact matrix. For example, for 10 kb data, each uniquely identified TAD boundary was flanked on either side by 10 kilobases for a total width around each boundary point of 20 kb. A genomic bin was labeled as having a TAD boundary within it (Y=1) if it overlapped with a particular flanked boundary. Otherwise, the bin was labeled as not containing a boundary (Y=0) (Figure 1).

@@

### FEATURE ENGINEERING AND REDUCTION USING GENOMIC ANNOTATIONS

@@

### RE-SAMPLING TECHNIQUES

In order to evaluate the problem of imbalanced classes, we then evaluated four re-sampling techniques (in addition to no re-sampling).

@@

2. **Random Under-Sampling (RUS)**: All of the minority classes from the training set were used. Sampling without replacement was used to obtain the same number of the majority classes. We performed 50 iterations of random under-sampling. At each iteration a random forest classification algorithm was performed. Performance metrics were aggregated by taking the average across all of the iterations.

3. **Random Over-Sampling (ROS)**: All of the majority classes from the training set were used. Sampling with replacement was used to obtain the same number of the minority classes. Similar to the under-sampling technique, 50 iterations were performed and the results of a random forest classification were aggregated by taking the average across iterations.

4. **SMOTE** (Synthetic Minority Over-Sampling Technique): This method incorporates both random under- and over-sampling. Under-sampling is performed without replacement from the majority class, while over-sampling is performed by creating new synthetic observations using the minority class [@chawla2002smote]. To account for the bias introduced by the under-sampling, we performed 50-iterations of SMOTE. We chose the parameters that would create perfectly balanced classes. The `DMwR` R package was used to perform SMOTE (version 0.4.1).

@@

### CLASSIFICATION & CROSS VALIDATION

A classification algorithm in the form of a random forest was performed on the various resampled and reduced training sets and validated using the same test set for all models in the ensemble. To reduce bias due to random dataset generation, 10-fold cross validation was used. The default number of features to consider at each node of the random forest algorithm was set as the square root of the number of features in the model. Likewise, the number of ensemble trees to aggregate was set at 500. Once the model was implemented, it was assessed using the test set, and performance metrics were then recorded.

### PERFORMANCE METRICS

The efficacy of different ensemble systems was compared using the F1-score and MCC. These metrics are defined as follows:

$$F1 - Score = \frac{2TP}{2TP + FP + FN}$$
$$MCC = \frac{TP \times TN - FP \times FN}{\sqrt{ (TP + FP)(TP+FN)(TN+FP)(TN+FN) }}$$

Here, TP refers to the number of bins correctly identified as containing a TAD boundary (true positives), FP refers to the number of bins incorrectly identified as containing a TAD boundary (false positives), TN refers to the number of bins correctly identified as not containing a TAD boundary (true negatives), and FN refers to the number of bins incorrectly identified as not containing a TAD boundary. Each of these quantities are obtained from the confusion matrix created by validating the model on the testing set. The evaluation metrics listed above are all considered single threshold measures. That is, they are defined for individual thresholds that separated class prediction. Thererfore, they are unable to give an overview of the potential range of performance as thresholds vary. In order to evaluate the trade-off between precision and recall, we also considered a threshold-free measure in form of the area under the precision recall curve (AUPRC). [??? This seem to be a good place to describe it in more details]

### MODEL ASSESSMENT AND PREDICTION

We first assessed predictor type performance using random forest classification algorithms by utilizing each set of predictor types as the feature space, in addition to a model with all types combined. Models were evaluated across each combination of resampling technique applied to the training set, for each cell line. Performance was based on the AUPRC. Recursive feature elimination (RFE) was then used to determine the optimal number of features for models with the best performing predictor type, across resampling techniques and cell lines. [??? Unclear at shich step this was done] In doing so, random forest models were again used to recursively assign weights to a set of features of a given size. For computational efficiency we evaluated feature subset sizes in descending order of powers of 2. The optimal number of features was determined by the AUPRC of the corresponding model. Optimal performance was assessed and the feature spaced was reduced accordingly.

For the task of identifying the functional genomic elements that were most predictive of the formation of TAD boundaries, we evaluated the variable importances associated with each predictor of the random forest models associated with each resampling technique. For each model and within each tree, the AUPRC on the out-of-bag portion of the data was recorded. Then the same was done after permuting each predictor variable. The difference between the two AUPRC values was then averaged over all trees, and normalized by the standard error [??? of what?]. All random forest algorithms were implemented in R using the `caret` package version 6.0 [@kuhn2012caret].

>>>>

Commonly used measures of model performance for binary classifiers include the area under the receiver operating curve (AUROC). The popularity surrounding the use of ROC curves is that they do not depend on a threshold. However, when the prevalence of an event is low--that is, a defined genomic region containing a TAD boundary--mean probabilities are biased towards the majority class. This results in the area of interest under an ROC curve to be compressed toward a small corner in the ROC space [@prati2011survey]. While not directly affecting the value of the AUROC, this can have severe consequences on the interpretability of the model's performance. A better threshold-free measure to use when dealing with imbalanced data is the area under the precision-recall curve (AUPRC). The precision-recall curve uses as its axes the precision and recall of the model. Both the precision and recall of a model include the true positives in the numerator of their respective ratios, thereby yielding more appropriate measures of a model's ability to accurately classify positive classes. Because of this, tt has been shown that the use of AUPRC with imbalanced datasets yields more accurate and intuitive interpretation of model performance when compared to AUROC [@saito2015precision]. Likewise, metrics like accuracy and specificity are also problematic when evaluating model performance on imbalance data due to being highly influenced by the correct classification of the majority class. Better measures to consider are the precision and recall themselves. Therefore, composite metrics that involve precision and recall, like the F1-Score, Youden's Index (YI), and Matthew's Correlation Coefficient (MCC) offer better interpretation, especially in the instance of imbalanced classes [@jeni2013facing]. The F1-Score can be interpreted as a weighted average of the precision and recall values, while Youden's index is interpreted as the arithmetic mean between sensitivity (recall) and specificity. MCC is, in essence, the correlation coefficient between the observed and predicted binary classifications. Each metric uses all 4 quadrants of the confusion matrix, thus providing more balanced measures in the case of imbalanced data [@chicco2017ten].

@@

To establish the data for modeling, the genome was binned into equally sized intervals defined by the data resolution for the corresponding contact matrix (Figure 1). For example, for domain data obtained from a 10 kb resolution contact matrix, the genome was binned into 10 kb intervals. For the boundary data, in order to accommodate for the uncertainty in exact boundary location, the boundary points were flanked on both sides by the size of the resolution of the corresponding contact matrix. For example, for 10 kb data, each uniquely identified TAD boundary was flanked on either side by 10 kilobases for a total width around each boundary point of 20 kb. A genomic bin was labeled as having a TAD boundary within it (Y=1) if it overlapped with a particular flanked boundary. Otherwise, the bin was labeled as not containing a boundary (Y=0).

@@

### FEATURE ENGINEERING AND RE-SAMPLING

@@

In order to assess the influence of imbalanced classes, we then evaluated four re-sampling techniques (in addition to no re-sampling), composed of data from each predictor type. Each re-sampling technique is described in detail below.

@@

2. **Random Under-Sampling (RUS)**: All of the minority classes from the training set were used. Sampling without replacement was used to obtain the same number of the majority classes.

3. **Random Over-Sampling (ROS)**: All of the majority classes from the training set were used. Sampling with replacement was used to obtain the same number of the minority classes.

4. **SMOTE** (Synthetic Minority Over-Sampling Technique): This method incorporates both random under- and over-sampling. Under-sampling is performed without replacement from the majority class, while over-sampling is performed by creating new synthetic observations using the minority class [@chawla2002smote]. The `DMwR` R package was used to perform SMOTE (version 0.4.1).

@@

To determine the best predictor type we performed separate random forest models on each set of predictor types and evaluated the precision-recall curves, as well as the subsequent AUPRC. To confirm the optimality of the feature engineering, this was performed for each combination of re-sampling technique, resolution, and cell line.

### FEATURE REDUCTION

Recursive feature elimination (RFE) was then used to determine the optimal number of features for models with the best performing predictor type. For each set of predictor type, random forest models were performed recursively on subsets of the feature space. The specific subsets of features were chosen as powers of 2 for computational efficiency. Model performance for each subset was evaluated using AUPRC. The optimal set of featues to include in downstream analyses was determined to be the subset in which the AUPRC leveled off. Again, to confirm that results were consistent, the reduction process was repeated for each combination of re-sampling technique, resolution, and cell line.

### FINAL MODEL ASSESSMENT AND PERFORMANCE

Once the appropriate predictor type and set of features were determined, a final set of random forest classification models were performed to fully compare re-sampling techniques. To reduce bias due to random dataset generation, 10-fold cross validation was used. The default number of features to consider at each node of the random forest algorithm was set as the square root of the number of features in the model. Likewise, the number of ensemble trees to aggregate was set at 500. All random forest algorithms were implemented in R using the `caret` package version 6.0 [@kuhn2012caret]. Once the model was implemented, it was assessed using the test set, and performance metrics were then recorded.

The efficacy of each re-sampling technique was compared using the F1-score, YI, and  MCC. These metrics are defined as follows:

$$F1 - Score = \frac{2TP}{2TP + FP + FN}$$

$$YI = \dfrac{TP}{TP+FN} + \dfrac{TN}{TN+FP} - 1$$

$$MCC = \frac{TP \times TN - FP \times FN}{\sqrt{ (TP + FP)(TP+FN)(TN+FP)(TN+FN) }}$$

Here, TP refers to the number of bins correctly identified as containing a TAD boundary (true positives), FP refers to the number of bins incorrectly identified as containing a TAD boundary (false positives), TN refers to the number of bins correctly identified as not containing a TAD boundary (true negatives), and FN refers to the number of bins incorrectly identified as not containing a TAD boundary. Each of these quantities are obtained from the confusion matrix created by validating the model on the testing set.

Lastly, for the task of identifying the genomic annotations that were most predictive of the formation of TAD boundaries, we evaluated the variable importances associated with each feature of the random forest model associated with the best combination of reduced predictor type and re-sampling technique, across each cell line. To calculate variable importances for each model, the AUPRC on the out-of-bag portion of the data within each tree was recorded. Then the same was done after permuting each predictor variable. The difference between the two AUPRC values was then averaged over all trees, and normalized by the standard error of the differences. All measures of importances were then rescaled to have a maximum value of 100 for better interpretation.
